---
title: "Data manipulation and visualization"
subtitle: "tidyverse: dplyr, tidyr, ggplot2"
date: 02.02.2017
font-family: 'Brill'
transition: none
output:
  slidy_presentation:
    df_print: paged
    footer: "Presentation link: https://goo.gl/fGi6Iv"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### 1.1 Data Types
* Vector
```{r}
c(42, 99, 43)
```
* Matrix, Array...
```{r}
matrix(1:12, nrow=3,ncol=4)
```
* List
```{r}
list(
  n_lectures = 12,
  l_topics = c("begining", "data manipulation", "descriptive stats"),
  h_topics = c("data manipulations", "descriptive stats")
  )
```

* Data Frame
```{r}
data.frame(
  names = c("Olya", "Ilya", "Sasha", "George"),
  lecturer = c(TRUE, TRUE, FALSE, TRUE),
  lecturer_experience = c(19, 6, 0, 3)
  )
```


[See Data Type Conversion page](http://www.statmethods.net/management/typeconversion.html)

### 1.2 Data Frame exploration
There are some embedded data frames (e. g. `mtcars`, `cars`, `iris`). How many rows and columns?
```{r}
nrow(iris) # returns the number of rows
ncol(mtcars) #  returns the number of columns
```
```{r}
head(cars) # returns the first 6 rows
head(cars, 4) # returns the first 4 rows
tail(cars) # returns the last 6 rows
summary(cars)  # produce some stats
str(cars)  # shows the structure: variables, their type
```

### 1.3 Data Frame Indexing
```{r}
mtcars$mpg # shows the mpg vector
mtcars[3,7] # shows the 3. row, 7. column
mtcars[3,] # shows the 3. row
mtcars[,7] # shows the 7. column
mtcars[mtcars$mpg < 20, ] # show all rows with the mpg value lower then 20
```

### 2. Tidyverse
The [tidyverse](https://blog.rstudio.org/2016/09/15/tidyverse-1-0-0/) is a set of packages:

* dplyr, for data manipulation
* ggplot2, for data visualisation
* tidyr, for data tidying
* readr, for data import
* purrr, for functional programming
* tibble, for tibbles, a modern re-imagining of data frames

Install tidyverse package using install.packages("tidyverse")

Load tidyverse package
```{r, message= FALSE}
library(tidyverse)
```
In this presentation is used the folowing version of tideverse:
```{r}
packageVersion("tidyverse")
packageVersion("dplyr")
packageVersion("ggplot2")
```

### 3.1 Join dataframe by row or column

```{r}
my_data_1 <- mtcars[5:20,] # select a subset of mtcars
my_data_2 <- mtcars[17:29,] # select a subset of mtcars
combine_rows <- rbind.data.frame(my_data_1, my_data_2)
nrow(my_data_1); nrow(my_data_2); nrow(combine_rows)
```

```{r}
my_data_3 <- mtcars[,3:7] # select a subset of mtcars
my_data_4 <- mtcars[,6:11] # select a subset of mtcars
combine_cols <- cbind.data.frame(my_data_3, my_data_4)
ncol(my_data_3); ncol(my_data_4); ncol(combine_cols)
```

### 3.2 Joins (dplyr)
```{r}
languages <- data.frame(
  languages = c("Selkup", "French", "Chukchi", "Kashubian"),
  countries = c("Russia", "France", "Russia", "Poland"),
  iso = c("sel", "fra", "ckt", "pol")
  )
languages
country_population <- data.frame(
  countries = c("Russia", "Poland", "Finland"),
  population_mln = c(143, 38, 5))
country_population
inner_join(languages, country_population)
left_join(languages, country_population)
right_join(languages, country_population)
anti_join(languages, country_population)
full_join(country_population, languages)
```

